\ \\ \LARGE\bf Speculative Evaluation in Particle Swarm Optimization

### 2010-04-20 honors_thesis/ppsn_paper/paper.tex
shortened more text; first person shooting

Particle swarm optimization (PSO) has previously been parallelized only by 
adding more particles to the swarm or by parallelizing the evaluation of the
objective function.  However, some functions are more efficiently optimized with more
iterations and fewer particles.  Accordingly, we take inspiration from 
speculative execution commonly performed in modern processors and propose what
we call speculative evaluation in PSO (SEPSO).  Future positions of the particles are
speculated and evaluated in parallel with current positions, performing two
iterations of PSO at once.

@@

In this paper we consider PSO parallelization strategies for clusters of
hundreds of processors and functions for which a single evaluation will take
at least seconds, but probably minutes.  Our purpose
is to explore the question of what to do with a hundred processors when 50 or
100 particles is the recommended swarm size, and simply increasing the swarm
size encounters diminishing returns.

To solve this problem, we propose a method for performing two iterations of PSO
at the same time in parallel that we call speculative evaluation.  The name
comes from an analogy to speculative execution (also known as branch
prediction) a technique commonly used in processors.  Modern processors, when
faced with a branch on which they must wait (e.g. a memory cache miss), guess
which way the branch will go and start executing, ensuring that any changes
can be undone.  If the processor guesses right, execution is
much farther ahead than if it had idly waited on the memory reference.  If
it guesses wrong, execution restarts where it would have been anyway.

In this paper we show that the results of standard PSO can be reproduced
\emph{exactly}, two iterations at a time, using a speculative approach adapted
from speculative execution. We prove that the standard PSO equation can be
factored such that a set of speculative positions can be found which will
\emph{always} include the position computed in the next iteration.  By
computing the value of the objective function for each of the speculative
positions at the same time the algorithm evaluates the objective function for
the current position, it is possible to know the objective function values for
both the current and the next iteration at the same time.  We demonstrate this
principle by implementation and show that it produces exactly the same results
as standard PSO, but two iterations at a time.  The resulting implementation
runs efficiently on large clusters where the number of processors is much
larger than a typical or reasonable number of particles, producing better
results in less ``wall-clock'' time.

@@

Particle swarm optimization was proposed in 1995 by James Kennedy and Russell
Eberhart~\cite{kennedy-1995-particle-swarm-optimization}.  The algorithm is
used to intelligently search a multi-dimensional space by mimicking the
swarming and flocking behavior of birds and other animals. It is a social
algorithm that depends on interaction between particles to quickly and
consistently approximate the optimal solution to a given objective function.

@@

Changing the way neighbors are defined (often called the ``topology'' in the
literature) has a significant effect on the performance of the algorithm.  In
the Ring topology, each particle has one neighbor to either side of it; in the
Complete topology, every particle is a neighbor to every other
particle~\cite{bratton-2007-defining-a-standard-for-pso}.  In all topologies a
particle is also a neighbor to itself in that its own position and value are
considered when updating the particle's neighborhood best, $\nbest$.  Thus with
$p$ particles, using the Ring topology each particle with index $i$ has three
neighbors: $i-1$, $i$ (itself), and $i+1$.  With the Complete topology, each
particle has $p$ neighbors.

@@

The PSO algorithm can be trivially parallelized by distributing the processing
needed for each particle on across processors.  But as we have seen in
\fig{evals-sphere}, for some functions, and for large numbers of
processors, just adding particles reaches a point of diminishing returns.
That is, adding processors does not help us reach any given level of fitness
appreciably faster.  Instead of adding particles we employ a speculative
evaluation approach that allows us to perform iterations two at a time.

To do so we must first refactor PSO such that the determination of the value
of the objective function is separate from the rest of the computation.  For
simplicity, this discussion will describe the case where PSO is performing
function minimization using the Ring topology.  In this example, each particle
has two neighbors, the ``right neighbor'' and ``left neighbor,'' whose
positions are represented as $\rightn$ and $\leftn$ respectively.  Though we
will only describe the case of the Ring topology, this method can easily be
extended to arbitrary topologies.

>>>>

Particle swarm optimization (PSO) has previously been parallelized only by 
adding more particles to the swarm or by parallelizing the evaluation of the
objective function.  However, some functions are more efficiently optimized with more
iterations and fewer particles.  Accordingly, we take inspiration from 
speculative execution performed in modern processors and propose
speculative evaluation in PSO (SEPSO).  Future positions of the particles are
speculated and evaluated in parallel with current positions, performing two
iterations of PSO at once.

@@

In this paper we consider PSO parallelization strategies for clusters of
hundreds of processors and functions for which a single evaluation will take
at least several seconds.  Our purpose
is to explore the question of what to do with hundreds of processors when 50 or
100 particles is the ideal swarm size, and simply adding particles
yields diminishing returns.

To solve this problem, we propose a method for performing two iterations of PSO
at the same time in parallel that we call speculative evaluation.  The name
comes from an analogy to speculative execution (also known as branch
prediction) a technique commonly used in processors.  Modern processors, when
faced with a branch on which they must wait (e.g., a memory cache miss), guess
which way the branch will go and start executing, ensuring that any changes
can be undone.  If the processor guesses right, execution is
much farther ahead than if it had idly waited on the memory reference.  If
it guesses wrong, execution restarts where it would have been anyway.

In this paper we show that the results of standard PSO can be reproduced
\emph{exactly}, two iterations at a time, using a speculative approach similar
to speculative execution. We prove that the standard PSO equation can be
factored such that a set of speculative positions can be found which will
\emph{always} include the position computed in the next iteration.  By
computing the value of the objective function for each of the speculative
positions at the same time the algorithm evaluates the objective function for
the current position, it is possible to know the objective function values for
both the current and the next iteration at the same time.
The resulting implementation
runs efficiently on large clusters where the number of processors is much
larger than a typical or reasonable number of particles, producing better
results in less ``wall-clock'' time.

@@

Particle swarm optimization was proposed in 1995 by James Kennedy and Russell
Eberhart~\cite{kennedy-1995-particle-swarm-optimization}.  This social
algorithm, inspired by the flocking behavior of birds, is used to quickly and
consistently find the global optimum of a given objective function in a
multi-dimensional space.

@@

Changing the way neighbors are defined, usually called the ``topology,''
has a significant effect on the performance of the algorithm.  In
the Ring topology, each particle has one neighbor to either side of it; in the
Complete topology, every particle is a neighbor to every other
particle~\cite{bratton-2007-defining-a-standard-for-pso}.  In all topologies a
particle is also a neighbor to itself in that its own position and value are
considered when updating the particle's neighborhood best, $\nbest$.  Thus with
$p$ particles, using the Ring topology each particle with index $i$ has three
neighbors: $i-1$, $i$ (itself), and $i+1$.  With the Complete topology, each
particle has $p$ neighbors.

@@

The PSO algorithm can be trivially parallelized by distributing the computation
needed for each particle across processors.  But
for some functions, as shown in \fig{evals-sphere},
adding particles yields diminishing returns.
That is, adding processors does not help reach any given level of fitness
appreciably faster.  Instead of adding particles, speculative
evaluation performs iterations two at a time.

Speculative evaluation is made possible by refactoring
PSO such that evaluating the
the objective function is separate from the rest of the computation.  For
simplicity, this discussion will describe the case where PSO is performing
function minimization using the Ring topology.  In this example, each particle
has two neighbors, the ``right neighbor'' and ``left neighbor,'' whose
positions are represented as $\rightn$ and $\leftn$ respectively.  Though we
will only describe the case of the Ring topology, this method is easily
extended to arbitrary topologies.
